---
title: "SAnjEmBhav"
output: html_document
date: "2024-03-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
socialvul <-read.csv("SVI_2018_US_county.csv")
```

## R Markdown

## Part 1 
##Introduction

##Social vulnerability refers to the level of social conditions that are associated with a community and its ability to react and adapt to challenges that it may face. This could be in the form of high poverty, diseases, social conditions, and so much more. We were especially interested in this dataset because it relates to topics like stress and resilience, important topics and skills for us as college students to consistently hone and develop.Important terms to know with regard to this dataset include the Social Vulnerability Index, which can serve as a factor that allows for the early identification of communities that may need more support in times of hazardous conditions. Counties in this dataset are divided into tracts, and 15 factors are measured for each of these tracts. Some examples include floods, disease, chemical exposure, etc.

## Part 2
## Written some code to explore and show your reader the structure of your dataset and the different variables involved, performing some initial data restructuring, filtering, and cleaning.

```{r cars}
socialvul <- read.csv("SVI_2018_US_county.csv")
library(dplyr)
str(socialvul)
head(socialvul)

# find the sum of E_POV after grouping them based on states
aggregate(socialvul$E_POV, list(socialvul$ST_ABBR), FUN = sum)

aggregate(socialvul$EP_UNEMP, list(socialvul$ST_ABBR), FUN = sum)

aggregate(socialvul$E_NOHSDP, list(socialvul$ST_ABBR), FUN = sum)
```


## Visualized the distribution of some of these variables and calculated some core tendency measures.
str()

## Put some thought into how you are breaking up your code into chunks that you will contextualize for your reader.

## Started writing some text walking your reader through what the chunks of code you've produced so far do, according to the conventions of literate programming.

## Including Plots

You can also embed plots, for example:

Code for plots
```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
